# CodeBook
## Coursera Data Science Specialization - Getting and Cleaning Data
Describes the variables, the data, and any transformations or work that you performed to clean up the data

## Raw data
The data can be automatically downloaded and unzip by the `get_data.R` script. Once extracted you will find 
a series of files under the `UCI HAR Dataset` directory, the `README.txt` file describes the content for each one
of the files in the directory.

This data set holds information about the accelerometer and gyroscope sensors of a Samsung phone as 30 subjects carried
on regular activities during the day. The data is divided into two sub-sets, one for training and one for testing, these
datasets were sampled randomly for the purposes of training and testing machine learning algorithms.

## Analysis
The analysis has a clear objective in mind as instructed by the assessment, it followed a phased approach to produce
the desired outcome in the form of a file that can be read by R containing the desired variables.

The first step of the analysis takes care of merging the individual subsets for training and testing. A simple
`rbind` was used to concatenate the two datasets together, worth noting that, because merging the training and
testing datasets was a reusable piece of functionallity, because there are other training and test files pertaining to
the same observation, it was implemented as a function that could be reused when merging training and test datasets
for other observed variables.

In the second step it finds the variables we are interested in the full range of feature variables available 
(there 561 available variables). The variables of interest are the mean and the standard deviation of the 
multitude of readings from the mobile sensors. The approach for finding the variables uses the `grep` function to 
identify the presence of either the word "mean" or "std" in the variable name to then use the indexes of where 
these character strings were found to select just the columns of interest from the merged dataSet create on the first phase.
The variable selection process brought down the number of variables to 81, all of which are properly described at the
bottom of this page.

In the third step, activity data for the training and test dataset are merged using the same function used in step 1 but with
different input data. Because the activity data is coded by a number for which activity names described are defined 
in the file `activity_labels.txt` that file is read and a new variable called `activityLabel` is created in the 
merged activity data frame that is a factor of the existing activity code leveled by the names defined in 
`activity_labels.txt`, in effect this creates a variable that displays as human readable description of the 
activity observed, e.g., walking, sitting, standing etc, but that is internally stored as a code that matches the 
original definitions found in the raw training and test acitivity datasets, e.g., waking is coded as activity 1.

Still in step three, to make sure that the new factor variable was equivalent to the original 
numeric coding of the raw data, I used a table to inspect the distribution of the factors and the original 
activity numeric codes as shown bellow

```{r echo=FALSE}
extractDir <- "./"
dataDir <- paste0(extractDir, "UCI HAR Dataset")
mergeDataSets <- function(type) {
  trainData <- read.table(paste0(dataDir, "/train/", type, "_train.txt"))
  testData <- read.table(paste0(dataDir, "/test/", type, "_test.txt"))
  
  # Checks if both datasets contain the same number of variables
  if (ncol(trainData) != ncol(testData)) stop("Something is wrong with the data")
  
  mergedData <- rbind(trainData, testData)
  
  if (nrow(mergedData) != nrow(trainData) + nrow(testData)) stop("Something is wrong with merge")
  
  mergedData
}
```

```{r}
mergedActData <- mergeDataSets("y")

names(mergedActData) <- "activity"

# Reads the activities descriptive labels
activityLabels <- read.table(paste0(dataDir, "/activity_labels.txt"))

mergedActData$activityLabel <- factor(mergedActData$activity, labels=activityLabels[,2])

# Checks the levels and labels, see if nothing was wronging coded into factors
table(mergedActData)
```

## Variables
```{r echo=FALSE}
vars <- read.csv("variables.txt", header = TRUE, check.names=FALSE)
kable(vars)
```